Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-22 paper/paper.tex
adding a reference to selective CPS. minor edit.

The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithms. \texttt{\$state}
is a Clojure hash map:
\begin{lstlisting}[style=default]
(def initial-state
  "initial program state"
  {:log-weight 0.0,
   :result nil,
   ::mem {},
   ::store nil,
   ... })
\end{lstlisting}
which records inference-relevant information under various keys
such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
The full list of map entries depends on the inference
algorithm. Except for transformation of
the \texttt{mem} form (which converts a function to one with memoization), 
CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  In
Clojure, checkpoints are represented by values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe}).
The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

@@

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program. Hence, \texttt{mem}'s
memory is stored as a nested dictionary in the program state
introduced during CPS transformation (function \texttt{mem-cps}).  
Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn [] (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply foo
        ;; memoize result in state
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state P)))))
\end{lstlisting}

@@

When an Anglican function is transformed into a Clojure
function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary
parameters are added to the beginning of the parameter list ---
continuation and state.  Correspondingly, when a function
\textit{call} is transformed\iftoggle{full}{ (by
\texttt{cps-of-application} or \texttt{cps-of-apply})}{}, the
current continuation and the state are passed to the called
function. Anglican can also call Clojure functions; however,
Clojure functions do not expect these auxiliary parameters.
To allow the mixing of Anglican (CPS-transformed) and Clojure
function calls in Anglican code, the Anglican compiler must be
able to recognize `primitive' (that is, implemented in Clojure
rather than in Anglican) functions.

>>>>

The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithms. \texttt{\$state}
is a Clojure hash map:
\begin{lstlisting}[style=default]
(def initial-state
  "initial program state"
  {:log-weight 0.0, ;; map :log-weight to 0.0
   :result nil,       ;; map :result to nil
   ::store nil,       ;; map :store to nil
   ::mem {},          ;; map ::mem to the empty map 
   ... })
\end{lstlisting}
which records inference-relevant information under various keys
such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
The full list of map entries depends on the inference
algorithm. Except for transformation of
the \texttt{mem} form (which converts a function to one with memoization), 
CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  
Checkpoints are values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
in Clojure. The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

@@

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument  and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program. Hence, \texttt{mem}'s
memory is stored as a nested dictionary in the program state
introduced during CPS transformation (function \texttt{mem-cps}).  
Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same argument. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn [] (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply foo
        ;; memoize result in state
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state P)))))
\end{lstlisting}

@@

When an Anglican function is transformed into a Clojure
function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary
parameters are added to the beginning of the parameter list ---
continuation and state.  Correspondingly, when a function
\textit{call} is transformed\iftoggle{full}{ (by
\texttt{cps-of-application} or \texttt{cps-of-apply})}{}, the
current continuation and the state are passed to the called
function. Anglican can also call Clojure functions; however,
Clojure functions do not expect these auxiliary parameters.
To allow the mixing of Anglican (CPS-transformed) and Clojure
function calls in Anglican code, the Anglican compiler must CPS-transform
expressions selectively~\cite{Nielsen01}. It must be
able to recognize `primitive' (that is, implemented in Clojure
rather than in Anglican) functions, and invoke
those functions in a direct, not CPS, style.
